General purpose gene annotation using deep learning

### 2020-01-09 paper/main.tex

We pre-processed and stored the genomic information by using Geenuff \citealp{Denton19}. Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. In the case of unambiguous input data, the genomic sequence was transformed into matrix that represents a one hot encoding with 4 rows. For the other case see Suplementary material .. (). The annotations are also transformed into a 4-row one hot encoding matrix, where a one in each of the rows represents one of the possible classes {\it Intergenic}, {\it UTR}, {\it Exon} or {\it Intron}. An non-coding intron can be inferred from the prediction order. A full description of all generated data arrays can be found in table \ref{Tab:01}.

@@

\subsection{Model Architecture and Usage}
We use a 4-layer deep stacked BLSTM network with 256 units per layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (citation). All relevant hyperparameters can be found in Table \ref{Tab:02} or in the Helixer source code repository.

@@

>>>>

We pre-processed and stored the genomic information by using Geenuff \citealp{Denton19}. Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. In the case of unambiguous input data, the genomic sequence was transformed into matrix that represents a one hot encoding with 4 rows. For the other case see Suplementary material .. (). The annotations are also transformed into a 4-row one hot encoding matrix, where a one in each of the rows represents one of the possible classes {\it Intergenic}, {\it UTR}, {\it CDS} or {\it Intron}. An non-coding intron can be inferred from the prediction order. A full description of all generated data arrays can be found in table \ref{Tab:01}.

The classes are not balanced equally and the tradeoff between the number of intergenic and intronic bases is quite different in both groups of genomes. Table \ref{Tab:statistics} shows the ratio of class occurences, together with other overall metrics.

\begin{table}[!t]
\processtable{Data group statistics\label{Tab:statistics}} {
\begin{tabular}{@{}lll@{}}
\toprule & EnsemblMetazoa & JGI Phytosome\\
\midrule
Fraction of classes  & 0.6, 0.02, 0.1, 0.2 & 0.6, 0.02, 0.1, 0.2 \\
(Intergenic, UTR, CDS, Intron) & & \\
Geenuff error rate & 0.1 (+- 0.02) & 0.3 (+- 0.02) \\
Average genome size in Gbp& 1.1 (+- 0.5) & 0.3 (+- 0.1) \\
Average gene length & 23000 (+- 5000) & 3370 (+- 50) \\
\botrule
\end{tabular}}{Values are averages across all genomes of a data group}
\end{table}

@@

Model evaluation during training was done with a metric that we call the {\it Genic F1}. This score is very similar to a classic F1 score calculated across multiple classes, except that it takes into account that the data is

\subsection{Model Architecture and Usage}
	We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization (citation ? ) between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (citation). All relevant hyperparameters can be found in Table \ref{Tab:02} or in the Helixer source code repository.

@@

\subsection{Evaluation of Augustus}
What Marvin did, why we don't use Genic F1 for comparison, etc
